---
title: "Beer Case Study"
author: "Paul Swenson, Yusi Zhang"
date: "2020-01-11"
output: html_document
---

# Brewery/Beer data analysis 
```{r}
library(ggplot2)
library(maps)
library(tidyverse)
library(mapproj)
library(doBy)
library(DataCombine)
library(class)
library(caret)
library(e1071)

#import data
beer = read.csv("../data/Beers.csv")
brewery = read.csv("../data/Breweries.csv")

#Note: US Population data was pulled from https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States_by_population on 1/17/2020
us_pop = read.csv("../data/US_Population.csv")

```

# 1. How many breweries are present in each state? 
Answer: AK AL AR AZ CA CO CT DE FL GA HI IA ID IL IN KS KY LA MA MD ME MI MN MO MS MT NC ND NE 
 7  3  2 11 39 47  8  2 15  7  4  5  5 18 22  3  4  5 23  7  9 32 12  9  2  9 19  1  5 
NH NJ NM NV NY OH OK OR PA RI SC SD TN TX UT VA VT WA WI WV WY 
 3  3  4  2 16 15  6 29 25  5  4  1  3 28  4 16 10 23 20  1  4 

```{r}


# remove Washington DC for plotting in the map
brewery2 = brewery[-c(228),]
brewery2$State <- factor(brewery2$State)

# Counts of breweries in each state
summary(brewery2$State)

lookup = data.frame(abb = state.abb, State = state.name) #makes a data frame with State name and abbreviation. 
colnames(brewery2)[4] = "abb" # Change Column Name

brewery_merged <- 
  left_join(brewery2, lookup, by = 'abb')

# plot on heatmap
brewery2 = merge(brewery2,lookup,"abb") # make one dataset with state names and abb
breweryMapData = count(brewery2,State) #count up the occurance of each state. 
colnames(breweryMapData)[2] = "Breweries" #change "n" to "Breweries"
breweryMapData$region <- tolower(breweryMapData$State)
breweryMapData2 = breweryMapData[-1]
states <- map_data("state")
map.df <- merge(states,breweryMapData2, by="region", all.x=T)
map.df <- map.df[order(map.df$order),]
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=Breweries))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+ggtitle("Breweries by State")+
coord_map()
```

# 2. Merge beer data with the breweries data. Print the first 6 observations and the last six observations to check the merged file. 
```{r}
#merge brew id between tables
#rename beer brew_id column to Brew_ID
colnames(beer)[5] = "Brew_ID" 

beer_brew <- 
  left_join(brewery, beer, by = 'Brew_ID')
colnames(beer_brew)[2] = "Brewery Name" # Change Column Name
colnames(beer_brew)[5] = "Beer Name" # Change Column Name
head(beer_brew, 6)
tail(beer_brew, 6)
```

# 3. Address the missing values in each column
ABV and IBU have 62 and 1005 missing values respectively
```{r}
# count the missing values in each column
colSums(is.na(beer_brew))

# ibu is right skewed, but almost half of the data are na values. This will be an issue.
hist(beer_brew$IBU)

# abv is normal with little skew, we are ok to use the mean to replace

# we will not be replacing the missing values since there are so few for abv
# we will not be replacing the missing values for IBU since there are so many
replace_abv = FALSE
hist(beer_brew$ABV)

if( replace_abv == TRUE )
{
  beer_brew$ABV[is.na(beer_brew$ABV)] <- with(beer_brew, ave(ABV, 
    FUN = function(x) mean(x, na.rm = TRUE)))[is.na(beer_brew$ABV)]
  sum(is.na(beer_brew$ABV))
  hist(beer_brew$ABV)
}
```

# 4. Compute the median alcohol content and international bitterness unit for each state. Plot a bar chart to compare.
```{r}
# drop rows without IBU values
beer_brew_dropped_IBU <- beer_brew %>% drop_na(IBU)

# get the median values and sort them high to low
beer_brew_dropped_IBU_sorted <- 
  beer_brew_dropped_IBU %>% 
   group_by(State) %>%
   summarise(median_ibu = median(IBU) )
  
beer_brew_dropped_IBU_sorted <- beer_brew_dropped_IBU_sorted[order(-beer_brew_dropped_IBU_sorted$median_ibu),]

##
## plot median IBU in map
##

colnames(beer_brew_dropped_IBU_sorted)[1] = "abb" # Change Column Name

# plot on heatmap
IBU_merged = merge(beer_brew_dropped_IBU_sorted,lookup,"abb") # make one dataset with state names and abb
colnames(IBU_merged)[2] = "Median_IBU" #change "n" to "Median IBU"
IBU_merged$region <- tolower(IBU_merged$State)
IBU_merged2 = IBU_merged[-1]
states <- map_data("state")
map.df <- merge(states,IBU_merged2, by="region", all.x=T)
map.df <- map.df[order(map.df$order),]
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=Median_IBU))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+ggtitle("Median IBU by State")+
coord_map()


# drop ABV rows without median
beer_brew_dropped_ABV <- beer_brew %>% drop_na(ABV)

# calculate the median values for abv then sort them high to low
beer_brew_dropped_ABV_sorted <- 
  beer_brew_dropped_ABV %>% 
   group_by(State) %>%
   summarise(median_abv = median(ABV) )
  
beer_brew_dropped_ABV_sorted <- beer_brew_dropped_ABV_sorted[order(-beer_brew_dropped_ABV_sorted$median_abv),]

##
## plot median abv
##

colnames(beer_brew_dropped_ABV_sorted)[1] = "abb" # Change Column Name

# Scale abv to a percentage value to increase readability
beer_brew_dropped_ABV_sorted$median_abv = beer_brew_dropped_ABV_sorted$median_abv * 100.0

# plot on heatmap
ABV_merged = merge(beer_brew_dropped_ABV_sorted,lookup,"abb") # make one dataset with state names and abb
colnames(ABV_merged)[2] = "Median_ABV" #change "n" to "Median ABV"
ABV_merged$region <- tolower(ABV_merged$State)
ABV_merged2 = ABV_merged[-1]
states <- map_data("state")
map.df <- merge(states,ABV_merged2, by="region", all.x=T)
map.df <- map.df[order(map.df$order),]
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=Median_ABV))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+ggtitle("Median ABV by State")+
coord_map()
```

# 5. Plot the maximum IBU and ABV in each state.
```{r}
# calculate the maximum IBU in each state
max_IBU <- beer_brew_dropped_IBU %>% 
   group_by(State) %>%
   summarise(mx = max(IBU) ) 

#Oregon has the max IBU at 138

##
## plot max ibu
##

colnames(max_IBU)[1] = "abb" # Change Column Name

max_IBU_merged = merge(max_IBU,lookup,"abb") # make one dataset with state names and abb
colnames(max_IBU_merged)[2] = "Max_IBU" #change "n" to "Max_IBU"
max_IBU_merged$region <- tolower(max_IBU_merged$State)
max_IBU_merged2 = max_IBU_merged[-1]
states <- map_data("state")
map.df <- merge(states,max_IBU_merged2, by="region", all.x=T)
map.df <- map.df[order(map.df$order),]
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=Max_IBU))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+ggtitle("Max IBU by State")+
coord_map()



# copy the dataset from earlier and scale ABV to a percentage value for readability
max_ABV <- beer_brew_dropped_ABV
max_ABV$ABV <- max_ABV$ABV * 100.0

# calculate the maximum value in each state
max_ABV <- max_ABV %>% 
   group_by(State) %>%
   summarise(mx = max(ABV) ) 

##
## plot max abv
##

colnames(max_ABV)[1] = "abb" # Change Column Name

max_ABV_merged = merge(max_ABV,lookup,"abb") # make one dataset with state names and abb
colnames(max_ABV_merged)[2] = "Max_ABV_Percentage" #change "n" to "Max_ABV"
max_ABV_merged$region <- tolower(max_ABV_merged$State)
max_ABV_merged2 = max_ABV_merged[-1]
states <- map_data("state")
map.df <- merge(states,max_ABV_merged2, by="region", all.x=T)
map.df <- map.df[order(map.df$order),]
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=Max_ABV_Percentage))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+ggtitle("Max ABV Percentage by State")+
coord_map()
```

# 6. Comment on the summary statistics and distribution of the ABV variable.
Beer ABV's are slightly right skewed (probably since people like to have stronger beers) and mostly normally distributed. There is a peak around 0.05 which could be the maximum beer strength to sell in specific states' grocery stores. For reference: at least colorado, Utah, and Minnesota have laws that restrict the abv to less than 5% for grocery store sales.
```{r}
# find basic statistics on the abv column
summary(beer_brew$ABV)

# plot it in a histogram
ggplot(beer_brew, aes(x=ABV))+
  geom_histogram(color="darkblue", fill="lightblue")+ggtitle("ABV Histogram")
```

# 7. Is there an apparent relationship between the bitterness of the beer and its alcoholic content?
```{r}
# generate scatter plot between IBU and ABV
scatter <- ggplot(beer, aes(IBU, ABV))
scatter + 
  ggtitle("Alcohol content by volume vs International bitterness unit") +
  geom_point() + 
  geom_smooth(method = "lm", colour = "Red", se = F) + 
  labs(x = "International bitterness unit", y = "Alcohol content by volume") 

# create a linear model of the relationship and print a summary about it
beer.lm = lm(ABV ~ IBU, data=beer) 
summary(beer.lm)

#based on the scatterplot for the data that contain both IBU and ABV, it looks like there is a moderate positive linear relation shipi between the two variables. r2= 0.4493 
```

# 8. Budweiser would also like to investigate the difference with respect to IBU and ABV between IPAs (India Pale Ales) and other types of Ale (any beer with “Ale” in its name other than IPA).  You decide to use KNN classification to investigate this relationship.  Provide statistical evidence one way or the other. You can of course assume your audience is comfortable with percentages … KNN is very easy to understand conceptually. 
```{r}
# find any styles that have " IPA" in them and save them to a new dataframe
ipa_beers <- grepl.sub(beer, " {0,1}\\({0,1}[iI][pP][aA]\\){0,1} {0,1}", c("Style"), keep.found = TRUE)
ipa_beers$Style = 'IPA'

# find any styles that have " ALE" in them and rename to ALE
ale_beers <- grepl.sub(beer, " {0,1}[aA][lL][eE] {0,1}", c("Style"), keep.found = TRUE)

# remove subset of beers that contain IPA in the style
ale_beers <- grepl.sub(ale_beers, " {0,1}\\({0,1}[iI][pP][aA]\\){0,1} {0,1}", c("Style"), keep.found = FALSE)
ale_beers$Style = 'ALE'

# 70% training data 30% test data
split_perc = 0.7

set.seed(6)

# combine the remaining ipa and ale beers to form the base dataset
knn_beers <- rbind(ipa_beers, ale_beers)

# select only the used columns
knn_beers <- knn_beers %>% select("ABV", "IBU", "Style")
knn_beers$IBU <- as.numeric(knn_beers$IBU)

# remove any NA values for IBU and ABV
knn_beers_dropmv <- knn_beers%>% drop_na(IBU,ABV)

# split the dataset into training and testing datasets 
trainIndices = sample(1:dim(knn_beers_dropmv)[1],round(split_perc * dim(knn_beers_dropmv)[1]))
train_beers = knn_beers_dropmv[trainIndices,]
test_beers = knn_beers_dropmv[-trainIndices,]

# typical k value should be sqrt of the number of samples in your dataset
k_value = ceiling(sqrt(length(train_beers$Style)))

#convert column to factor 
train_beers$Style <- as.factor(train_beers$Style)

# create scatterplot with labels for ales and ipas
knn_beers_dropmv %>% ggplot(aes(x=IBU, ABV, color=Style)) +
  geom_point()+ggtitle("ABV vs IBU Classification")

# create knn model and evaluate the fit
classifications = knn(train_beers[,c(1,2)],test_beers[,c(1,2)],train_beers$Style, prob = TRUE, k = k_value)
class_table <- table(classifications,test_beers$Style)
confusionMatrix(class_table)
```

# 9. Scale breweries per state by proportion of population to overall US population then look at their least produced beers (ABV and IBU)
```{r}
# change the state values to lower case
us_pop$region <- tolower(us_pop$State)

# join the us population data with the brewery count data from before
brewery_pop <- left_join(breweryMapData2, us_pop, by="region")

# scale the beweries to match the proportion of the united states population
brewery_pop$Breweries_Scaled <- brewery_pop$Percent * brewery_pop$Breweries

# plot the scaled brewery counts
states <- map_data("state")
map.df <- merge(states,brewery_pop, by="region", all.x=T)
map.df <- map.df[order(map.df$order),]
ggplot(map.df, aes(x=long,y=lat,group=group))+
  geom_polygon(aes(fill=Breweries_Scaled))+
  geom_path()+ 
  scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90")+ggtitle("Breweries Per Capita")+
coord_map()
```